---
title: "Figures and tables for the main manuscript of 'Dissecting depression symptoms: multi-omics clustering uncovers immune-related subgroups and cell-type specific dysregulation'"
author: "Jonas Hagenberg et al."
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, include = FALSE}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggpubr)
library(readxl)
library(FactoMineR)
library(factoextra)
library(ggalluvial)
library(openxlsx)

source("figures_and_tables/plotting_functions.R")
source("figures_and_tables/test_functions.R")
source("figures_and_tables/reporting_functions.R")
```

# Data 4 cluster solution age/BMI, cytokines, RNAseq
```{r}
general_data <- read_in_general_data(proband_type = "current_cases_new_cidi",
                                     cytokine_version = "additional_correct",
                                     healthy_control_version = "strict",
                                     add_lifetime_cases = FALSE)

gsva_go_gene_sets <- GSEABase::getGmt("c5.go.bp.c5.go.mf.v2022.1.Hs.entrez.gmt")


define_cluster_colours <- function() {
  c("MIDS" = "#3fd482", "HIRDS1" = "#ffca4d",
    "LIRDS" = "#6857d9", "HIRDS2" = "#fa5c4b")
}

set_defaults <- function(
    compare_means_test = "kruskal.test",
    add_lifetime_cases = FALSE,
    add_healthy_controls = TRUE
) {
  list(
    compare_means_test = compare_means_test,
    add_lifetime_cases = add_lifetime_cases,
    add_healthy_controls = add_healthy_controls
  )
}

# phenotypes, cytokines, RNAseq
res_sumo_4_pheno_cyt_rna <- read.table("01_res_phenotypes_cytokines_rna_seed_max_iter_10000/k4/clusters.tsv",
                       header = TRUE)
res_sumo_4_pheno_cyt_rna$label <- res_sumo_4_pheno_cyt_rna$label + 1
res_sumo_4_pheno_cyt_rna <- res_sumo_4_pheno_cyt_rna %>% 
  rename(ID = sample, cluster = label)

# geneset data
gsva_go_uncorrected <- readRDS("gsva_go_uncorrected.rds")

# gene VI data
important_genes_initial_clustering <- readRDS("phenotypes_cytokines_rna_4_important_genes.rds")
```

# Data 3 cluster solution age/BMI/celltypes, cytokines, RNAseq
```{r}
define_cluster_colours_celltypes_v2 <- function() {
  c("reMIDS" = "#94d4b1", "reIDS" = "#b3b3b3",
    "reHIRDS" = "#ffe099")
}

res_sumo_3_pheno_cell_cyt_rna <- read.table("01_res_phenotypes_celltypes_14_cytokines_rna_seed_max_iter_10000/k3/clusters.tsv",
                       header = TRUE)
res_sumo_3_pheno_cell_cyt_rna$label <- res_sumo_3_pheno_cell_cyt_rna$label + 1
res_sumo_3_pheno_cell_cyt_rna <- res_sumo_3_pheno_cell_cyt_rna %>% 
  rename(ID = sample, cluster = label)
```

# Figure 1
The rest of the figure is designed in Illustrator.
```{r}
bmi_figure_1 <- generate_plot(
  type = "BMI",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  compare_means_test = NA,
  median_BMI_annotation = FALSE,
  x_axis_legend = element_blank(),
  cluster_names = data.frame(
    cluster = 1:4,
    cluster_new = c("MIDS", "HIRDS1", "LIRDS", "HIRDS2")
  ),
  cluster_colours = c("MIDS" = "#3fd482", "HIRDS1" = "#ffca4d",
    "LIRDS" = "#6857d9", "HIRDS2" = "#fa5c4b")
  ) +
  theme_classic() +
  theme(text = element_text(size = 36),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        line = element_line(size = 1)
        ) +
  labs(x = NULL, y = NULL)

bmi_figure_1$layers[[1]]$aes_params$size <- 1
bmi_figure_1$layers[[2]]$aes_params$size <- 1
bmi_figure_1$layers[[4]]$aes_params$size <- 1
```

```{r}
ggsave("../../../paper/figures/figure_1/violinplots_4.pdf",
       bmi_figure_1,
       width = 20, height = 20, units = "cm",
       scale = 1)
```

# Figure 2
Needs to be put together via Illustrator.
```{r}
heatmap_cases_legend_4_sumo_paper <- generate_plot(
  type = "heatmap",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  annotation_legend = TRUE,
  var_name_study_instead_med = TRUE,
  pretty_names = TRUE,
  show_psychopharmaca = FALSE,
  cytokine_version = "additional_unimputed",
  treeheight_row = 0,
  included_annotations = c("cluster", "study"),
  show_cytokine_panel_names = TRUE,
  sort_by_panel_names = TRUE,
  fontsize = 24
)
```
```{r}
ggsave("../../../paper/figures/figure_2/heatmap.pdf",
       heatmap_cases_legend_4_sumo_paper$gtable,
       width = 60, height = 40, units = "cm",
       scale = 1)
```

```{r}
variable_importance_4 <- generate_plot(
  type = "variable_importance",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  omic_type = "cytokines_RNA",
  rna_version = "standard",
  cytokine_version = "additional_unimputed",
  show_gene_names = TRUE,
  include_age_bmi = TRUE,
  show_annotation_string = FALSE,
  show_gene_id = FALSE
  ) +
  theme(text = element_text(size = 36),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank()
        )

variable_importance_4
```

```{r}
ggsave("../../../paper/figures/figure_2/variable_importance_broad.pdf",
       variable_importance_4,
       width = 60, height = 30, units = "cm",
       scale = 1)
```

Get the legend:
```{r}
variable_importance_4_with_legend <- generate_plot(
  type = "variable_importance",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  omic_type = "cytokines_RNA",
  rna_version = "standard",
  cytokine_version = "additional_unimputed",
  show_gene_names = TRUE,
  include_age_bmi = TRUE,
  show_annotation_string = FALSE,
  show_gene_id = FALSE,
  legend_position = "right"
  ) +
  theme(text = element_text(size = 36)
        )

legend_variable_importance_4 <- get_legend(variable_importance_4_with_legend)
as_ggplot(legend_variable_importance_4)
```

```{r}
ggsave("../../../paper/figures/figure_2/variable_importance_legend.pdf",
       legend_variable_importance_4,
       width = 50, height = 30, units = "cm",
       scale = 1)
```

Get the report to have the names variable importance values:
```{r}
values_variable_importance_4 <- generate_report(
  type = "variable_importance",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  omic_type = "cytokines_RNA",
  rna_version = "standard",
  cytokine_version = "additional_unimputed",
  show_gene_names = TRUE,
  include_age_bmi = TRUE
)

head(values_variable_importance_4, n = 40)
```

```{r}
age_4_v2 <- generate_plot(
  type = "age",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  compare_means_test = NA,
  x_axis_legend = element_blank(),
  cluster_names = data.frame(
    cluster = 1:4,
    cluster_new = c("MIDS", "HIRDS1", "LIRDS", "HIRDS2")
  ),
  cluster_colours = c("MIDS" = "#3fd482", "HIRDS1" = "#ffca4d",
    "LIRDS" = "#6857d9", "HIRDS2" = "#fa5c4b")
  ) +
  theme(text = element_text(size = 36),
        axis.text.x = element_text(angle = 45, hjust = 1)
        )

age_4_v2
```

```{r}
bmi_4_v2 <- generate_plot(
  type = "BMI",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  compare_means_test = NA,
  median_BMI_annotation = FALSE,
  use_unimputed_data = TRUE,
  x_axis_legend = element_blank(),
  cluster_names = data.frame(
    cluster = 1:4,
    cluster_new = c("MIDS", "HIRDS1", "LIRDS", "HIRDS2")
  ),
  cluster_colours = c("MIDS" = "#3fd482", "HIRDS1" = "#ffca4d",
    "LIRDS" = "#6857d9", "HIRDS2" = "#fa5c4b")
  ) +
  theme(text = element_text(size = 36),
        axis.text.x = element_text(angle = 45, hjust = 1)
        )

bmi_4_v2
```

Combine age & BMI:
```{r}
age_bmi_4_v2 <- ggarrange(
  age_4_v2,
  bmi_4_v2,
  ncol = 3,
  legend = "none"
)
```

```{r}
ggsave("../../../paper/figures/figure_2/age_bmi_broad_unimputed_v2.pdf",
       age_bmi_4_v2,
       width = 60, height = 30, units = "cm",
       scale = 1)
```

Calculate the mean BMI values per group:
```{r}
general_data$cytokine_uncorrected %>% 
  right_join(res_sumo_4_pheno_cyt_rna, by = "ID") %>% 
  group_by(cluster) %>% 
  summarise(mean_BMI = mean(BMI, na.rm = TRUE),
            median_BMI = median(BMI, na.rm = TRUE))
```

# Figure 3
Needs to be put together via Illustrator.
```{r}
rel_bmi_crp <- general_data$cytokine_uncorrected %>% 
  right_join(res_sumo_4_pheno_cyt_rna, by = "ID") %>% 
  mutate(cluster = case_when(
    cluster == 1 ~ "MIDS",
    cluster == 2 ~ "HIRDS1",
    cluster == 3 ~ "LIRDS",
    cluster == 4 ~ "HIRDS2"
  )) %>% 
  ggplot() +
  aes(x = BMI, y = hsCRP, col = cluster) +
  geom_point() +
  geom_smooth(size = 2) +
  theme_bw() +
  scale_colour_manual(values = define_cluster_colours()) +
  theme(
    text = element_text(size = 36)
  ) +
  coord_cartesian(ylim = c(-3, 4))

rel_age_crp <- general_data$cytokine_uncorrected %>% 
  right_join(res_sumo_4_pheno_cyt_rna, by = "ID") %>% 
  mutate(cluster = case_when(
    cluster == 1 ~ "MIDS",
    cluster == 2 ~ "HIRDS1",
    cluster == 3 ~ "LIRDS",
    cluster == 4 ~ "HIRDS2"
  )) %>% 
  ggplot() +
  aes(x = age, y = hsCRP, col = cluster) +
  geom_point() +
  geom_smooth(size = 2) +
  theme_bw() +
  scale_colour_manual(values = define_cluster_colours()) +
  theme(
    text = element_text(size = 36)
  ) +
  coord_cartesian(ylim = c(-3, 4))

rel_age_bmi <- general_data$cytokine_uncorrected %>% 
  right_join(res_sumo_4_pheno_cyt_rna, by = "ID") %>% 
  mutate(cluster = case_when(
    cluster == 1 ~ "MIDS",
    cluster == 2 ~ "HIRDS1",
    cluster == 3 ~ "LIRDS",
    cluster == 4 ~ "HIRDS2"
  )) %>% 
  ggplot() +
  aes(x = age, y = BMI, col = cluster) +
  geom_point() +
  geom_smooth(size = 2) +
  theme_bw() +
  scale_colour_manual(values = define_cluster_colours()) +
  theme(
    text = element_text(size = 36)
  )

rel_bmi_crp_age <- ggarrange(
  rel_bmi_crp,
  rel_age_crp,
  rel_age_bmi,
  nrow = 1,
  common.legend = TRUE,
  legend = "bottom",
  labels = "AUTO",
  font.label = list(size = 36))
```

```{r}
bdi_4 <- generate_plot(
  type = "BDI",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  compare_means_test = NA,
  x_axis_legend = element_blank()
  ) +
  theme(
    text = element_text(size = 36),
    axis.text.x = element_text(angle = 45, hjust = 1)
    )

bdi_4
```

```{r}
pretty_cluster_assignments_v2 <- get_cluster_assignments(
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  general_data = general_data,
  cluster_names = data.frame(
    cluster = 1:4,
    cluster_new = c("MIDS", "HIRDS1", "LIRDS", "HIRDS2")
  )
)
pretty_cluster_assignments_v2 <- pretty_cluster_assignments_v2[["cluster_assignments_df"]]
```


```{r}
sdnn_median_4 <- general_data$ecg %>% 
  left_join(pretty_cluster_assignments_v2, by = "ID") %>% 
  filter(!is.na(cluster)) %>% 
  ggplot(aes(x = cluster, y = SDNN)) +
  geom_violin(aes(fill = cluster)) +
  geom_boxplot(aes(group = cluster), width = 0.2) +
  theme_bw() +
  scale_fill_manual(values = c(define_cluster_colours(), "controls" = "#FFFFFF")) +
  theme(text = element_text(size = 36),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank()
  )

sdnn_median_4 <- add_dashes(sdnn_median_4, general_data$ecg %>% 
  left_join(pretty_cluster_assignments, by = "ID") %>% 
    filter(!is.na(cluster)))
sdnn_median_4
```

```{r}
lateral_ventricle_4_v1 <- generate_plot(
  type = "imaging_residuals",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  compare_means_test = NA,
  cluster_names = data.frame(
    cluster = 1:4,
    cluster_new = c("MIDS", "HIRDS1", "LIRDS", "HIRDS2")
  ),
  cluster_colours = c("MIDS" = "#3fd482", "HIRDS1" = "#ffca4d",
    "LIRDS" = "#6857d9", "HIRDS2" = "#fa5c4b"),
  remove_controls = TRUE,
  show_n = FALSE
  ) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
        )

lateral_ventricle_4_v2 <- lateral_ventricle_4_v1 +
  theme(text = element_text(size = 36))
```


Bring everything together:
```{r}
bdi_ecg_imaging_4 <- ggarrange(
  bdi_4,
  sdnn_median_4,
  lateral_ventricle_4_v2,
  nrow = 1,
  legend = "none",
  labels = c("D", "E", "F"),
  font.label = list(size = 36))

figure_3_v3 <- ggarrange(
  rel_bmi_crp_age,
  bdi_ecg_imaging_4,
  nrow = 2
)
```

```{r}
ggsave("../../../paper/figures/figure_3/figure_3_v3.pdf",
       figure_3_v3,
       width = 70, height = 60, units = "cm",
       scale = 1)
```

Calculate the Pearson correlation within the groups for the different analyses:
```{r}
general_data$cytokine_uncorrected %>% 
  right_join(res_sumo_4_pheno_cyt_rna, by = "ID") %>% 
  group_by(cluster) %>% 
  summarise(corr = cor(BMI, hsCRP, use = "pairwise.complete.obs", method = "pearson"),
            p_value = cor.test(BMI, hsCRP, use = "pairwise.complete.obs", method = "pearson")$p.value)
```

```{r}
general_data$cytokine_uncorrected %>% 
  right_join(res_sumo_4_pheno_cyt_rna, by = "ID") %>% 
  group_by(cluster) %>% 
  summarise(corr = cor(age, hsCRP, use = "pairwise.complete.obs", method = "pearson"),
            p_value = cor.test(age, hsCRP, use = "pairwise.complete.obs", method = "pearson")$p.value)
```

```{r}
general_data$cytokine_uncorrected %>% 
  right_join(res_sumo_4_pheno_cyt_rna, by = "ID") %>% 
  group_by(cluster) %>% 
  summarise(corr = cor(age, BMI, use = "pairwise.complete.obs", method = "pearson"),
            p_value = cor.test(age, BMI, use = "pairwise.complete.obs", method = "pearson")$p.value)
```

# Figure 4
Needs to be put together via Illustrator.
```{r}
cytokine_boxplots_4 <- generate_plot(
  type = "cytokines",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  cytokine_list = c("IL_1RA", "hsCRP", "PlGF", "MCP_1", "TNF_alpha",
                    "MIP_1beta", "MCP_4"),
  violin = TRUE,
  complete_names = FALSE,
  cytokine_version = "additional_unimputed",
  sort_by_list = TRUE,
  pretty_names = TRUE,
  ncol = 4
  ) +
  theme(
    axis.text.x = element_blank(),
    text = element_text(size = 36),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  ylab("normalised concentration")

cytokine_boxplots_4
```

```{r}
genes_boxplot_4 <- generate_plot(
    type = "RNA",
    general_data = general_data,
    cluster_assignments = res_sumo_4_pheno_cyt_rna,
    rna_type = "standard",
    gene_list = c("ENSG00000091129", "ENSG00000186462", "ENSG00000163520",
                  "ENSG00000155893", "ENSG00000172985", "ENSG00000158813",
                  "ENSG00000182489"),
    compare_means_test = NA,
    sort_by_list = TRUE,
    ncol = 4,
    facet_scale = "fixed",
    show_gene_id = FALSE
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 36),
    axis.title.x = element_blank()
  )

genes_boxplot_4
```

```{r}
cytokine_rna_boxplots_4 <- ggarrange(
  cytokine_boxplots_4,
  genes_boxplot_4,
  ncol = 1,
  common.legend = TRUE,
  legend = "none")

cytokine_rna_boxplots_4
```

```{r}
ggsave("../../../paper/figures/figure_4/boxplots_cytokines_genes_4.pdf",
       cytokine_rna_boxplots_4,
       width = 35, height = 70, units = "cm",
       scale = 1)
```

Create a heatmap of the top genesets:

```{r}
gsva_go_clean_uncorrected <- as.data.frame(t(gsva_go_uncorrected))
gsva_go_clean_uncorrected$ID <- rownames(gsva_go_clean_uncorrected)
rownames(gsva_go_clean_uncorrected) <- NULL

geneset_importance_go_uncorrected <- generate_report(
  type = "variable_importance",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  omic_type = "genesets",
  geneset_results = gsva_go_clean_uncorrected,
  include_p_value = TRUE
)

geneset_heatmap_go_uncorrected <- generate_plot(
  type = "heatmap_genesets",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  scale_expression = TRUE,
  # colour_palette = colorRampPalette(c("blue", "yellow"))(100),
  colour_palette = viridis::viridis(100),
  treeheight_row = 0,
  show_current_depression = FALSE,
  show_sex = FALSE,
  show_age = FALSE,
  show_crp = FALSE,
  show_psychopharmaca = FALSE,
  show_medication = FALSE,
  geneset_data = gsva_go_clean_uncorrected,
  geneset_list = geneset_importance_go_uncorrected %>% dplyr::slice(1:20) %>% pull(geneset),
  cluster_rows = FALSE,
  fontsize = 24
)
```
```{r}
ggsave("../../../paper/figures/figure_4/geneset_heatmap.pdf",
       geneset_heatmap_go_uncorrected,
       width = 60, height = 50, units = "cm",
       scale = 1)
```

```{r}
geneset_importance_go_uncorrected
```

Determine the number of genes in the gene sets:
```{r}
gsva_input_data_uncorrected <- general_data$rna %>% 
  dplyr::rename(ensembl_id = gene) %>% 
  left_join(mapping_info %>% dplyr::select(entrez_id, ensembl_gene_id),
          by = c("ensembl_id" = "ensembl_gene_id")) %>% 
  filter(!is.na(expression), !is.na(entrez_id)) %>% 
  dplyr::select(-c(ensembl_id)) %>% 
  pivot_wider(
    id_cols = entrez_id,
    names_from = ID,
    values_from = expression
  )

gene_names_uncorrected <- gsva_input_data_uncorrected$entrez_id
gsva_input_data_uncorrected <- as.matrix(gsva_input_data_uncorrected %>% dplyr::select(-entrez_id))
rownames(gsva_input_data_uncorrected) <- gene_names_uncorrected
```

```{r}
gene_sets_of_interest <- c(
  "GOBP_ANTEROGRADE_AXONAL_TRANSPORT",
  "GOBP_REGULATION_OF_MACROPHAGE_ACTIVATION",
  "GOBP_RESPONSE_TO_MINERALOCORTICOID",
  "GOBP_REGULATION_OF_CALCIUM_ION_TRANSMEMBRANE_TRANSPORTER_ACTIVITY",
  "GOBP_RESPONSE_TO_MOLECULE_OF_BACTERIAL_ORIGIN",
  "GOBP_REGULATION_OF_SECRETION",
  "GOBP_POSITIVE_REGULATION_OF_TRANSMEMBRANE_TRANSPORT",
  "GOBP_POSITIVE_REGULATION_OF_NEURON_DEATH"
)
genes_in_gene_set <- lapply(gene_sets_of_interest, function(one_set) {
  GSEABase::geneIds(gsva_go_gene_sets[one_set])
})

genes_intersected_with_data_set <- lapply(genes_in_gene_set, function(one_set) {
  set_data <- data.frame(one_set)
  colnames(set_data) <- "gene"
  genes_in_data_set <- data.frame(gene = dimnames(gsva_input_data_uncorrected)[[1]])
  res <- set_data %>% 
    inner_join(genes_in_data_set, by = "gene") %>% 
    nrow()
  names(res) <- names(one_set)
  res
})
```

Gene correlations
```{r}
genes_correlation_4 <- generate_plot(
  type = "RNA_correlation",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  rna_type = "standard",
  gene_list = c("ENSG00000091129", "ENSG00000186462", "ENSG00000163520",
                "ENSG00000155893", "ENSG00000172985", "ENSG00000158813",
                "ENSG00000182489"),
  tl.cex = NULL,
  sample_type = "cases"
) +
  viridis::scale_fill_viridis(name = "Corr", limits = c(-0.45, 0.65),
                              option = "magma") +
  theme(
    text = element_text(size = 36),
    legend.key.width = unit(4.5, "lines")
  )

genes_correlation_4  
```

```{r}
ggsave("../../../paper/figures/figure_4/correlation_genes_4.pdf",
       genes_correlation_4  ,
       width = 20, height = 20, units = "cm",
       scale = 1)
```

```{r}
cytokines_correlation_4 <- generate_plot(
  type = "cytokine_correlation",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  cytokine_version = "additional_unimputed",
  cytokine_list = c("IL_1RA", "hsCRP", "PlGF", "MCP_1", "TNF_alpha",
                    "MIP_1beta", "MCP_4"),
  tl.cex = NULL,
  pretty_names = TRUE,
  sample_type = "cases"
) +
  viridis::scale_fill_viridis(name = "Corr", limits = c(-0.45, 0.65),
                              option = "magma") +
  theme(
    text = element_text(size = 36),
    legend.key.width = unit(4.5, "lines")
  )

cytokines_correlation_4  
```

Combine the correlation plots
```{r}
correlation_plots_4 <- ggarrange(
  genes_correlation_4,
  cytokines_correlation_4,
  ncol = 2,
  common.legend = TRUE,
  legend = "bottom"
)

correlation_plots_4
```
```{r}
ggsave("../../../paper/figures/figure_4/correlation_genes_cytokines_4.pdf",
       correlation_plots_4  ,
       width = 40, height = 20, units = "cm",
       scale = 1)
```

Give the values of correlation:
```{r}
cyt_corr <- calculate_cytokine_correlation(
  data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  cytokine_version = "additional_unimputed",
  sample_type = "cases",
  pretty_names = TRUE,
  cytokine_list = c("IL_1RA", "hsCRP", "PlGF", "MCP_1", "TNF_alpha",
                    "MIP_1beta", "MCP_4")
)

cyt_corr
```

```{r}
rna_corr <- calculate_RNA_correlation(
  data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  rna_type = "standard",
  gene_list = c("ENSG00000091129", "ENSG00000186462", "ENSG00000163520",
                "ENSG00000155893", "ENSG00000172985", "ENSG00000158813",
                "ENSG00000182489"),
  sample_type = "cases"
)

rna_corr
```

Calculate mean absolute correlations:
```{r}
mean_cyt_corr <- lapply(seq_len(ncol(cyt_corr)), function(col) {
  # the matrix is symmetric, so I can use the col value to exclude the row
  # where the correlation with itself is stored
  res <- mean(abs(cyt_corr[-col, col]))
  names(res) <- dimnames(cyt_corr)[[2]][col]
  res
})
unlist(mean_cyt_corr)
mean(unlist(mean_cyt_corr))

mean_rna_corr <- lapply(seq_len(ncol(rna_corr)), function(col) {
  # the matrix is symmetric, so I can use the col value to exclude the row
  # where the correlation with itself is stored
  res <- mean(abs(rna_corr[-col, col]))
  names(res) <- dimnames(rna_corr)[[2]][col]
  res
})
unlist(mean_rna_corr)
mean(unlist(mean_rna_corr))
```


# Figure 5
Alluvial plot between clusters:
```{r}
cluster_colours_celltypes_alluvial_v2 <- c(
  "MIDS" = "#3fd482",
  "HIRDS1" = "#ffca4d",
  "LIRDS" = "#6857d9",
  "HIRDS2" = "#fa5c4b",
  "reMIDS" = "#94d4b1",
  "reIDS" = "#b3b3b3",
  "reHIRDS" = "#ffe099"
  )

# prepare the data
data_alluvial_celltypes_v2 <- res_sumo_4_pheno_cyt_rna %>% 
  left_join(data.frame(
    cluster = 1:4,
    initial_clustering = c("MIDS", "HIRDS1", "LIRDS", "HIRDS2")
  ), by = "cluster") %>% 
  select(-cluster, `initial clustering` = initial_clustering) %>% 
  left_join(res_sumo_3_pheno_cell_cyt_rna, by = "ID") %>% 
  # rename(`including celltypes` = cluster) %>% 
  left_join(data.frame(
            cluster = 1:3,
            including_celltypes = c( "reMIDS", "reIDS", "reHIRDS")),
            by = "cluster") %>% 
  rename(`including celltypes` = including_celltypes) %>% 
  select(-cluster) %>% 
  pivot_longer(
    cols = -ID,
    names_to = "clustering",
    values_to = "cluster assignment"
  ) %>% 
  mutate(
    clustering = factor(clustering, levels = c("initial clustering",
                                               "including celltypes")),
    `cluster assignment` = factor(`cluster assignment`,
                                  levels = c("MIDS", "HIRDS1", "LIRDS", "HIRDS2",
                                             "reMIDS", "reIDS", "reHIRDS"))
    )

plot_alluvial_uncorrected_celltypes_v2 <- data_alluvial_celltypes_v2 %>% 
  ggplot() +
  aes(
    x = clustering,
    stratum = `cluster assignment`,
    alluvium = ID,
    fill = `cluster assignment`
  ) +
  geom_flow(stat = "alluvium", lode.guidance = "frontback") +
  geom_stratum() +
  geom_label(stat = "stratum",
             aes(label = after_stat(stratum)),
             fill = "white",
             # the theme size is in pt, the geom_label size is in mm
             size = 36 * 0.36) +
  scale_fill_manual(values = cluster_colours_celltypes_alluvial_v2) +
  scale_x_discrete(expand = c(0.1, 0.2)) +
  theme_bw() +
  theme(
    legend.position = "none",
    text = element_text(size = 36),
    axis.title.x = element_blank()
  )

plot_alluvial_uncorrected_celltypes_v2
```

```{r}
ggsave("../../../paper/figures/figure_5/alluvial_cluster_assignment_v2.pdf",
       plot_alluvial_uncorrected_celltypes_v2,
       width = 25, height = 30, units = "cm",
       scale = 1)
```

Calculate the percentages of old cluster assignments in new clusters:
```{r}
group_size_3_pheno_cell_cyt_rna <- res_sumo_3_pheno_cell_cyt_rna %>% 
  group_by(cluster) %>% 
  summarise(group_size = n())

res_sumo_4_pheno_cyt_rna %>% 
  rename(`uncorrected clustering` = cluster) %>% 
  left_join(res_sumo_3_pheno_cell_cyt_rna, by = "ID") %>%
  group_by(cluster, `uncorrected clustering`) %>% 
  summarise(n = n()) %>% 
  left_join(group_size_3_pheno_cell_cyt_rna, by = "cluster") %>%
  mutate(fraction = n / group_size)
```

Variable importance plot:
```{r}
variable_importance_3_celltypes <- generate_plot(
  type = "variable_importance",
  general_data = general_data,
  cluster_assignments = res_sumo_3_pheno_cell_cyt_rna,
  omic_type = "cytokines_RNA",
  rna_version = "standard",
  cytokine_version = "additional_unimputed",
  show_gene_names = TRUE,
  include_age_bmi = TRUE,
  include_celltypes = TRUE,
  force_cytokines_in_plot = TRUE,
  show_annotation_string = FALSE,
  show_gene_id = FALSE
  ) +
  theme(text = element_text(size = 36),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank()
        )

variable_importance_3_celltypes
```

```{r}
ggsave("../../../paper/figures/figure_5/variable_importance.pdf",
       variable_importance_3_celltypes,
       width = 50, height = 30, units = "cm",
       scale = 1)
```

Get the report to have the names variable importance values:
```{r}
values_variable_importance_3_celltypes <- generate_report(
  type = "variable_importance",
  general_data = general_data,
  cluster_assignments = res_sumo_3_pheno_cell_cyt_rna,
  omic_type = "cytokines_RNA",
  rna_version = "standard",
  cytokine_version = "additional_unimputed",
  show_gene_names = TRUE,
  include_age_bmi = TRUE,
  include_celltypes = TRUE
)

values_variable_importance_3_celltypes %>% 
  filter(variable %in% c("age", "BMI", "hsCRP", "IP_10",
                         "T cells CD4 memory resting",
                         "Dendritic cells activated",
                         "Monocytes",
                         "ZEB2 ENSG00000169554",
                         "LINC01550 ENSG00000246223"))
```

Distribution of celltypes:
```{r}
boxplot_celltypes_3_v2 <- generate_plot(
  type = "celltypes",
  general_data = general_data,
  cluster_assignments = res_sumo_3_pheno_cell_cyt_rna,
  celltype_type = "RNA_updated_14",
  cluster_colours = define_cluster_colours_celltypes_v2(),
  cluster_names = data.frame(
    cluster = 1:3,
    cluster_new = c("reMIDS", "reIDS", "reHIRDS")
  ),
  compare_means_test = NA
  ) +
  ylab("fraction") +
  theme(
    text = element_text(size = 36),
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

boxplot_celltypes_3_v2
```
```{r}
ggsave("../../../paper/figures/figure_5/celltypes_distribution_v2.pdf",
       boxplot_celltypes_3_v2,
       width = 50, height = 30, units = "cm",
       scale = 1)
```

Calculate the mean values across groups:
```{r}
general_data$celltype_data_rna_updated_14 %>% 
  pivot_longer(
    cols = -ID,
    names_to = "celltype",
    values_to = "percentage"
  ) %>% 
  right_join(res_sumo_3_pheno_cell_cyt_rna, by = "ID") %>% 
  group_by(celltype, cluster) %>% 
  summarise(mean_fraction = mean(percentage))
```


Percentages of different cell types (total):
```{r}
celltypes_total <- general_data$celltype_data_rna_updated_14 %>% 
  right_join(res_sumo_3_pheno_cell_cyt_rna, by = "ID") %>% 
  summarise(across(`B cells memory`:`T cells gamma delta`, median, na.rm = TRUE)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "fraction"
  ) %>% 
  dplyr::mutate(fraction = fraction / 100) %>% 
  ggplot() +
  aes(
    x = 1,
    y = fraction,
    fill = celltype
    ) +
  geom_col() +
  theme_bw() +
  scale_fill_manual(values = c("#5F4690","#1D6996","#38A6A5","#0F8554","#73AF48","#EDAD08","#E17C05","#CC503E", "#cc3d59", "#94346E","#6F4070","#994E95","#666666", "#000000")) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    text = element_text(size = 36)
  )
```

```{r}
ggsave("../../../paper/figures/figure_5/celltypes_total.pdf",
       celltypes_total,
       width = 30, height = 30, units = "cm",
       scale = 1)
```

```{r}
general_data$celltype_data_rna_updated_14 %>% 
  right_join(res_sumo_3_pheno_cell_cyt_rna, by = "ID") %>% 
  summarise(across(`B cells memory`:`T cells gamma delta`, median, na.rm = TRUE)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "celltype",
    values_to = "fraction"
  ) %>% 
  dplyr::mutate(fraction = fraction / 100) %>% 
  arrange(desc(fraction))
```


Geneset enrichment plot:
```{r}
geneset_heatmap_go_uncorrected_celltypes_v2 <- generate_plot(
  type = "heatmap_genesets",
  general_data = general_data,
  cluster_assignments = res_sumo_3_pheno_cell_cyt_rna,
  scale_expression = TRUE,
  colour_palette = viridis::viridis(100),
  cluster_colours = define_cluster_colours_celltypes_v2(),
  cluster_names = data.frame(
    cluster = 1:3,
    cluster_new = c("reMIDS", "reIDS", "reHIRDS")
  ),
  treeheight_row = 0,
  show_current_depression = FALSE,
  show_sex = FALSE,
  show_age = FALSE,
  show_crp = FALSE,
  show_psychopharmaca = FALSE,
  show_medication = FALSE,
  geneset_data = gsva_go_clean_uncorrected,
  geneset_list = geneset_importance_go_uncorrected_celltypes %>% dplyr::slice(1:20) %>% pull(geneset),
  cluster_rows = FALSE,
  fontsize = 24
)
```
```{r}
ggsave("../../../paper/figures/figure_5/geneset_heatmap_v2.pdf",
       geneset_heatmap_go_uncorrected_celltypes_v2,
       width = 60, height = 50, units = "cm",
       scale = 1)
```

```{r}
geneset_importance_go_uncorrected_celltypes
```

# Figure 6
```{r}
vegfa_boxplot_4 <- generate_plot(
  type = "cytokines",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  cytokine_list = c("VEGF_A_HS"),
  violin = TRUE,
  complete_names = FALSE,
  cytokine_version = "additional_unimputed",
  sort_by_list = TRUE,
  pretty_names = TRUE,
  ncol = 4
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 36),
    axis.title.x = element_blank()
  ) +
  ylab("normalised concentration")

vegfa_boxplot_4
```
```{r}
serpinf1_boxplot_4 <- generate_plot(
    type = "RNA",
    general_data = general_data,
    cluster_assignments = res_sumo_4_pheno_cyt_rna,
    rna_type = "standard",
    gene_list = c("ENSG00000132386"),
    compare_means_test = NA,
    sort_by_list = TRUE,
    ncol = 4,
    facet_scale = "fixed",
    show_gene_id = FALSE
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 36),
    axis.title.x = element_blank()
  )

serpinf1_boxplot_4
```

```{r}
vegfa_rna_boxplot_4 <- generate_plot(
    type = "RNA",
    general_data = general_data,
    cluster_assignments = res_sumo_4_pheno_cyt_rna,
    rna_type = "standard",
    gene_list = c("ENSG00000112715"),
    compare_means_test = NA,
    sort_by_list = TRUE,
    ncol = 4,
    facet_scale = "fixed",
    show_gene_id = FALSE
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 36),
    axis.title.x = element_blank()
  )

vegfa_rna_boxplot_4
```


```{r}
serpinf1_vegfa_vegfa_rna <- ggarrange(
  serpinf1_boxplot_4,
  vegfa_boxplot_4,
  vegfa_rna_boxplot_4,
  nrow = 1,
  common.legend = TRUE,
  legend = "bottom",
  labels = NULL,
  font.label = list(size = 36))

serpinf1_vegfa_vegfa_rna
```

```{r}
ggsave("../../../paper/figures/figure_6/serpinf1_vegfa_rna_boxplots.pdf",
       serpinf1_vegfa_vegfa_rna,
       width = 75, height = 30, units = "cm",
       scale = 1)
```

The UMAPs were generated in python, see the file `01_b_single_cell_figure_6.ipynb`.

# Table 1
Generate a table with information about age, sex, BMI, BDI, MDD/dysthymia
diagnosis, medication.

```{r}
id_data_with_status <- bind_rows(
  res_sumo_4_pheno_cyt_rna %>% select(ID) %>% mutate(status = "case"),
  general_data$control_ids %>% mutate(status = "control")
)

bmi_data <- general_data$cytokine_uncorrected %>% 
  select(ID, BMI)

sex_data <- general_data$imputed_data %>% 
  filter(.imp == 1) %>% 
  select(ID, sex)

mdd_diagnosis <- general_data$cidi_single %>% 
  mutate(has_depression_dysthymia = Depression_full_curr == 1 |
             Dysthymia_full_curr == 1 | Depression_subthr_curr == 1) %>% 
  select(ID, has_depression_dysthymia)

# get all the data together
data_sample_table <- general_data$age %>% 
  right_join(id_data_with_status, by = "ID") %>% 
  left_join(general_data$bdi_sumscores %>% select(ID, BDI_total), by = "ID") %>% 
  left_join(bmi_data, by = "ID") %>% 
  left_join(sex_data, by = "ID") %>% 
  left_join(mdd_diagnosis, by = "ID") %>% 
  mutate(has_depression_dysthymia = if_else(status == "control", FALSE, has_depression_dysthymia)) %>% 
  left_join(general_data$medication_data, by = "ID")
```

```{r}
data_sample_table_summary <- data_sample_table %>% 
  group_by(status) %>% 
  summarise(n = n(),
            across(c(age, BMI, BDI_total), .fns = list(mean = mean, sd = sd)),
            across(c(has_depression_dysthymia, t0_any_psych_medication),
                   .fns = list(fraction = ~sum(.x, na.rm = TRUE) / n(), n = ~sum(.x, na.rm = TRUE))),
            female_fraction = sum(sex == "female") / n(), female_n = sum(sex == "female")) %>% 
  mutate(age = paste0(round(age_mean, 1), " (", round(age_sd, 1), ")"),
         female = paste0(round(female_fraction * 100, 0), "% (", female_n, ")"),
         BMI = paste0(round(BMI_mean, 1), " (", round(BMI_sd, 1), ")"),
         `MDD or dysthymia diagnosis` = paste0(round(has_depression_dysthymia_fraction * 100, 0), "% (", has_depression_dysthymia_n, ")"),
         BDI = paste0(round(BDI_total_mean, 1), " (", round(BDI_total_sd, 1), ")"),
         `psychotropic drugs` = paste0(round(t0_any_psych_medication_fraction * 100, 0), "% (", t0_any_psych_medication_n, ")")) %>% 
  select(status, n, age, female, BMI, `MDD or dysthymia diagnosis`, BDI, `psychotropic drugs`)

data_sample_table_summary
```

Paste the following into Excel and then into the document as table:
```{r}
clipr::write_clip(data_sample_table_summary)
```

