---
title: "Statistical tests for 'Dissecting depression symptoms: multi-omics clustering uncovers immune-related subgroups and cell-type specific dysregulation'"
author: "Jonas Hagenberg et al."
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, include = FALSE}
library(dplyr)
library(tidyr)
library(stringr)

source("figures_and_tables/plotting_functions.R")
source("figures_and_tables/test_functions.R")
```

# Data
## Initial clustering
```{r}
general_data <- read_in_general_data(proband_type = "current_cases_new_cidi",
                                     cytokine_version = "additional_correct",
                                     healthy_control_version = "strict",
                                     add_lifetime_cases = FALSE)

set_defaults <- function(
    compare_means_test = "kruskal.test",
    add_lifetime_cases = FALSE,
    add_healthy_controls = TRUE
) {
  list(
    compare_means_test = compare_means_test,
    add_lifetime_cases = add_lifetime_cases,
    add_healthy_controls = add_healthy_controls
  )
}

# phenotypes, cytokines, RNAseq
res_sumo_4_pheno_cyt_rna <- read.table("01_res_phenotypes_cytokines_rna_seed_max_iter_10000/k4/clusters.tsv",
                       header = TRUE)
res_sumo_4_pheno_cyt_rna$label <- res_sumo_4_pheno_cyt_rna$label + 1
res_sumo_4_pheno_cyt_rna <- res_sumo_4_pheno_cyt_rna %>% 
  rename(ID = sample, cluster = label)
```

## Clustering corrected for age/sex/BMI
```{r}
res_sumo_3_cyt_rna_cov <- read.table("04_cytokines_rna_corrected_covariates/01_res_cytokines_rna_corrected_covariates_1_seed_max_iter_10000/k3/clusters.tsv",
                       header = TRUE)
res_sumo_3_cyt_rna_cov$label <- res_sumo_3_cyt_rna_cov$label + 1
res_sumo_3_cyt_rna_cov <- res_sumo_3_cyt_rna_cov %>% 
  rename(ID = sample, cluster = label)
```

## Reclustering with celltypes
```{r}
res_sumo_3_pheno_cell_cyt_rna <- read.table("01_res_phenotypes_celltypes_14_cytokines_rna_seed_max_iter_10000/k3/clusters.tsv",
                       header = TRUE)
res_sumo_3_pheno_cell_cyt_rna$label <- res_sumo_3_pheno_cell_cyt_rna$label + 1
res_sumo_3_pheno_cell_cyt_rna <- res_sumo_3_pheno_cell_cyt_rna %>% 
  rename(ID = sample, cluster = label)
```

# Statistics
Test the distribution of several phenotypes across the clusters.

## Initial clustering
BDI/MADRS
```{r}
generate_test(
  type = "BDI",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  method = "Tukey"
)
```

```{r}
generate_test(
  type = "MADRS",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  method = "Tukey"
)
```

```{r}
generate_test(
  type = "BDI_single",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna
)
```

CIDI diagnoses:
```{r}
generate_test(
  type = "cidi_diagnosis",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna
)
```

```{r}
generate_test(
  type = "cidi_depression_dysthymia",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna
)
```

Correlation of CRP, age and BMI
```{r}
data_correlation_initial_clustering <- general_data$cytokine_uncorrected %>% 
  right_join(res_sumo_4_pheno_cyt_rna, by = "ID")

# BMI & CRP
data_correlation_initial_clustering %>% 
  group_by(cluster) %>% 
  summarise(
    cor = cor.test(BMI, hsCRP, use = "complete.obs")$estimate,
    p.value = cor.test(BMI, hsCRP, use = "complete.obs")$p.value
  )

# age & CRP
data_correlation_initial_clustering %>% 
  group_by(cluster) %>% 
  summarise(
    cor = cor.test(age, hsCRP, use = "complete.obs")$estimate,
    p.value = cor.test(age, hsCRP, use = "complete.obs")$p.value
  )

# age & BMI
data_correlation_initial_clustering %>% 
  group_by(cluster) %>% 
  summarise(
    cor = cor.test(age, BMI, use = "complete.obs")$estimate,
    p.value = cor.test(age, BMI, use = "complete.obs")$p.value
  )
```

Distribution of Become and Optima cohorts across clusters
```{r}
data_studies_initial_clustering <- res_sumo_4_pheno_cyt_rna %>% 
  mutate(
    study = if_else(str_detect(ID, "^F"), "Become", "Optima")
  )

fisher.test(
  x = data_studies_initial_clustering$cluster,
  y = data_studies_initial_clustering$study
)
```
Exclude the MIDS cluster:
```{r}
data_studies_initial_clustering_wo_mids <- res_sumo_4_pheno_cyt_rna %>% 
  mutate(
    study = if_else(str_detect(ID, "^F"), "Become", "Optima")
  ) %>% 
  filter(cluster != 1)

fisher.test(
  x = data_studies_initial_clustering_wo_mids$cluster,
  y = data_studies_initial_clustering_wo_mids$study
)
```

ECG
```{r}
generate_test(
  type = "ECG",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  # with_exclusions means that the participants with low quality data were excluded
  QC_filtering = "with_exclusions"
)
```

Imaging
```{r}
test_imaging_thickness_ancova <- generate_test(
  type = "imaging_ancova",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  variable_type = "thickness_revised",
  model_type = "without_age2"
)

test_imaging_volume_ancova <- generate_test(
  type = "imaging_ancova",
  general_data = general_data,
  cluster_assignments = res_sumo_4_pheno_cyt_rna,
  variable_type = "volumina",
  model_type = "without_age2"
)

results_4_cluster_final_version <- bind_rows(
  test_imaging_thickness_ancova,
  test_imaging_volume_ancova %>% filter(variable %in% c("LatVent", "hippo"))
) %>% 
  filter(str_detect(term, "^cluster")) %>% 
  mutate(
    p_value_adjusted = p.adjust(p.value, method = "BH")
  )
```

## Clustering corrected for age/sex/BMI
```{r}
generate_test(
  type = "BDI",
  general_data = general_data,
  cluster_assignments = res_sumo_3_cyt_rna_cov,
  method = "Tukey"
)
```
```{r}
generate_test(
  type = "MADRS",
  general_data = general_data,
  cluster_assignments = res_sumo_3_cyt_rna_cov,
  method = "Tukey"
)
```

```{r}
generate_test(
  type = "age",
  general_data = general_data,
  cluster_assignments = res_sumo_3_cyt_rna_cov,
  method = "Tukey"
)
```

```{r}
generate_test(
  type = "BMI",
  general_data = general_data,
  cluster_assignments = res_sumo_3_cyt_rna_cov,
  method = "Tukey",
  bmi_source = "with_missing"
)
```

## Reclustering with cell types
```{r}
generate_test(
  type = "BDI",
  general_data = general_data,
  cluster_assignments = res_sumo_3_pheno_cell_cyt_rna,
  method = "Tukey"
)
```
```{r}
generate_test(
  type = "MADRS",
  general_data = general_data,
  cluster_assignments = res_sumo_3_pheno_cell_cyt_rna,
  method = "Tukey"
)
```